---
title: "Introduction to the 'h5ad' package"
author:
  - name: Martin Morgan
    affiliation:
    - Roswell Park Comprehensive Cancer Center
    email: mtmorgan.bioc@gmail.com
output:
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
date: "`r doc_date()`"
package: "`r pkg_ver('h5ad')`"
vignette: >
  %\VignetteIndexEntry{Introduction to the h5ad package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    crop = NULL
)
```

# Basics

## Install `h5ad`

This package is only available on GitHub. Install with

```{r install, eval = FALSE}
BiocManager::install("mtmorgan/h5ad")
```

## Citing `h5ad`

Please cite this package as follows:

```{r citation}
citation("h5ad")
```

# Quick start

Retrieve an `.h5ad` dataset from the [CELLxGENE][] repository.

[CELLxGENE]: https://cellxgene.cziscience.com/datasets

```{r "cxg-dataset", message=FALSE}
library(cellxgenedp)

dataset <-
    datasets() |>
    dplyr::filter(dataset_id == "65badd7a-9262-4fd1-9ce2-eb5dc0ca8039") |>
    select(dataset_id)

heart_h5ad_file <-
    left_join(dataset, files(), by = "dataset_id")  |>
    filter(filetype == "H5AD") |>
    files_download(dry = FALSE)
```

Create an `h5ad()` object by referencing the h5ad file. This object
summarizes the content of the file without loading data into memory.

```{r h5ad, message = FALSE}
library("h5ad")

h5ad <- h5ad(heart_h5ad_file)
h5ad
```

Extract the PCA or UMAP embeddings from the file.

- Use `j =` to extract only some dimensions of the embedding, as
  illustrated for PCA. The file includes 50 principle components, we
  extract the first three.
- Use `with =` to extract additional row- or column annotations
  associated with each row of the embedding. The examples below
  extract the `cell_type` classification of each cell, so that points
  can be colored by cell type.

```{r embedding}
pca <-
    h5ad |>
    column_embedding("X_pca", j = 1:3, with = c("donor_id", "cell_type"))

umap <-
    h5ad |>
    column_embedding("X_umap", with = "cell_type") |>
    slice(sample(n()))

umap |>
    count(cell_type, sort = TRUE)
```

Use base *R* functionality for static visualization.

```{r plot}
plot(X_umap_2 ~ X_umap_1, umap, pch = ".", col = cell_type)
```

Interactively visualize the data using `plotly`.

```{r plotly, message = FALSE}
library(plotly)
```

```{r plotly-umap}
plot_ly(
    umap,
    x = ~ X_umap_1, y =  ~ X_umap_2,
    color = ~ cell_type,
    colors = "Paired",
    type = "scatter", mode = "markers", opacity = 0.5,
    marker = list(size = 2L)
) |>
    ## `toWebGL()` **greatly** improves rendering performance!
    toWebGL()
```

```{r plotly-pca-3d}
plot_ly(
    pca |> slice(sample(n())),
    x = ~ X_pca_1, y = ~ X_pca_2, z = ~ X_pca_3,
    color = ~ cell_type,
    colors = "Paired",
    type = "scatter3d", mode = "markers", opacity = 0.5,
    marker = list(size = 1L)
) |> toWebGL()
```

# Reproducibility

Create the vignette.

```{r create-vignette, eval=FALSE}
rmarkdown::render("introduction-to-h5ad.Rmd", "BiocStyle::html_document")
```

Extract *R* code.

```{r extract-r-code, eval = FALSE}
knitr::knit("introduction-to-h5ad.Rmd", tangle = TRUE)
```

Session information (packages and versions used for producing this
vignette).

```{r reproduce3, echo=FALSE}
sessionInfo()
```
